### 2024-05-16 Chapter5.tex
Adding minor corrections from viva/annotated thesis

\chapter[Application of models to HSI data]{Application of analytical tissue models to hyperspectral imaging data}\label{chap:HSImodel}
\begin{center}
\begin{minipage}[b]{0.9\linewidth}
\small
\textbf{Foreword\,}
This chapter is as yet unpublished. 
\newline
AB has conducted all of the research presented in this chapter under the supervision of JS, MB, and TV. The gelatin-based phantom images were annotated by OM. 
\end{minipage}
\end{center}

@@

\section{Introduction}
As described in Chapter~\ref{chap:intro}, hyperspectral imaging (HSI) is able to capture spatially resolved, diffuse reflectance spectra via non-contact imaging~\citep{Lu2014,Giannoni2018,Calin2014,Shapey2019}, which could enable quantitative, spatially-resolved, tissue oxygenation ($StO_2$) extraction. Many methods have been considered for this ranging from two or three wavelength approaches~\citep{MacKenzie2018} to wavelength-dependent analytical models including those investigated in Chapters~\ref{chap:1layer} and \ref{chap:2layer} as these have been some of the prominent methods of doing so~\citep{Yudovsky2009, Jacques1999, Clancy2015, Clancy2020}. Other methods include deep-learning based approaches, however these rely on models or simulation to generate their training data~\citep{Wirkert2016}, or models based on the derivatives of the spectra, however these rely on extensive calibration and assume that the scattering contribution is wavelength-independent which may not be true for all tissues~\citep{Holmer2018}.

@@

In this chapter, we quantify the impact of reducing the spectral resolution and introduction of noise on the performance of the analytical models investigated in Chapters~\ref{chap:1layer} and \ref{chap:2layer} by simulating a snapshot camera response from the Monte Carlo datasets used in these previous chapters. These impacts are then evaluated with experimental HSI data of the phantoms detailed in Chapter~\ref{chap:1layer}. Finally, the models are applied to neurosurgical HSI data obtained intra-operatively as part of the HELICoiD dataset~\cite{Fabelo2019}. % and NHSI datasets.

\section{Methods}
\subsection{Camera response simulation}\label{sec:MCcameras}
To obtain datasets of simulated data with known ground truth parameters, the datasets described in Sections~\ref{sec:methodsMC} and \ref{sec:methodreference2} of internal and skin tissues are used. For each spectrum in the dataset, the camera response is simulated by convolving the band responses of the snapshot HSI camera (depicted in Figures~\ref{fig:rulerspectrum} and \ref{fig:simulatingcam}) and the integral of this is divided by the band response integral to account for differing band quantum efficiencies as defined in Equation~\eqref{eq:simulatingcam}. 
\begin{equation}
	c_{n} = \frac{\int r(\lambda)b_n(\lambda) d\lambda}{\int b_n(\lambda) d\lambda}
\label{eq:simulatingcam}
\end{equation}
Here $c_n$ is the measured value from band $n$ by the camera when the true spectrum is $r(\lambda)$ and the band response for this band is $b_n(\lambda)$. Each band response has a major peak around a central wavelength which is the wavelength associated with that band in the simulated spectrum. This is similar to the approach taken in Chapter~\ref{chap:SWB} for the ruler reflectivity correction as it takes into account both the true diffuse reflectance spectrum of the sample and the band responses of the camera used to capture it which do not perfectly sample single wavelengths. Random Gaussian noise can also be added at this stage where the mean is taken to be 0 and varying standard deviations ($\sigma$) are added. $c_n$ is then corrected using an correction matrix to account for the parasitic cross-talk effects as in~\ref{methodnecessity}~\citep{Pichette2017}. Whilst the illuminant spectrum will impact the measured spectrum, we employ appropriate white balancing as described in Chapter~\ref{chap:SWB} which should take account of this as well as the reflectivity of the white reference so the illuminant spectrum is not included in this simulation. An example of the stages of this process can be seen in Figure~\ref{fig:simulatingcam}. This allows a variety of datasets to be simulated with the same set of ground truth parameters for both quantitative and mean-normalised relative data. Two noise levels are investigated alongside noiseless simulations. The maximum noise level chosen ($\sigma = 3pp$ before mean normalisation) is similar to the standard deviation size seen in the highly controlled dataset investigated in Chapter~\ref{chap:SWB} in units of percentage point (pp).  
\begin{figure}[h!]
    \centering 
    \begin{subfigure}{0.7\textwidth}
    \centering
        \includegraphics[width=\textwidth]{simulatingcam.pdf}
        \caption{}
        \label{fig:simulatingcamsubfigure}
    \end{subfigure}
    \begin{subfigure}{0.62\textwidth}
        \centering
        \includegraphics[width=\textwidth]{band_responses.pdf}
        \caption{}
        \label{fig:bandresponses}
    \end{subfigure}
    % \includegraphics[width=0.7\textwidth]{simulatingcam.pdf}
    \caption{Top: An example of the stages of converting a Monte Carlo diffuse reflectance spectrum (\textcolor{blue}{blue solid}) to a simulated camera response (\textcolor{red}{red $*$}) via the band response convolution (\textcolor{purple}{purple $+$}) and cross-talk correction. Bottom: A depiction of the band responses used in this camera simulation.}
    \label{fig:simulatingcam}
\end{figure}

@@

    \caption{Table displaying the percentages of dyes [acid red 1 (AR1), acid red 14 (AR14), crystal violet (CV)] a used for each dye configuration alongside the total dye scaled concentration in arbitrary units. The intralipid (v/v \%) imaged for each configuration are also indicated.}

@@

A white reference of a Spectralon tile of 95\% is taken under the same lighting conditions and imaging plane as the phantoms. This white reference is either used directly for white balancing a frame of the hyperspectral video to produce quantitative data or to generate spectral sensitivities used to construct mean normalised relative data as in Chapter~\ref{chap:SWB}. This demonstrates both the highest fidelity and more realistic data processing pathways. Following this, bilinear demosaicing is used and cross-talk correction is applied per pixel of the frame also detailed in Chapter~\ref{chap:SWB}. At this stage, it is possible to temporally average frames to improve the signal to noise ratio of the resulting hypercube. Finally, any pixel spectra containing values above 99.5\% or below 0.5\% are discarded to reduce the impact of specular reflections and to ensure the data is within the dynamic range of the sensor. 

@@

\subsection{Neurosurgical HSI dataset}\label{sec:NeuroHSIdata}
The HELICoiD dataset is leveraged in this work which captures data in the wavelength range 400-1000nm using a pushbroom spatial scanning HSI camera~\citep{Fabelo2019}. This data is obtained as part of a collaboration between four universities, three industrial partners, and two hospitals in the European Union~\citep{Fabelo2019}. This provides 36 high spectral resolution images of 22 patients which are annotated with use of spectral angle mapper to identify four key classes: normal tissue, tumour tissue, blood vessel, and background elements". This data is white balanced using associated flatfield white references included in the dataset. The resulting spectra are then normalised to reduce the impact of any differences in relative position of the camera to the data and white reference as discussed in Chapter~\ref{chap:SWB}. For this work we consider only the spectral range of 450-600nm as this is the range of use for the analytical models considered. 
%Secondly, a NeuroHSI dataset is considered which is obtained using the same snapshot mosaic camera (Ximea utilising the IMEC CMV2K-SSM4X4-VIS sensor, Germany) as in Section~\ref{sec:imagingphantoms} whose bands are within the 450-600nm wavelength range. In contrast to Section~\ref{sec:imagingphantoms}, this is mounted to a neurosurgical microscope (Zeiss Kinevo 900, Germany) whose integrated light source and lenses are used for simpler integration into the neurosurgical workflow, allowing for use in a wide variety of neurosurgical cases. 
%%This light source and optical system was not fully characterised and therefore the vignetting profile cannot be assumed to follow the trends seen in Chapter~\ref{chap:SWB} and so the full white balancing algorithm using the sterile ruler cannot be implemented for quantitative data. In it's place, 
%In each imaging timepoint of this study, the position of the neurosurgical microscope is recorded and used to obtain a flatfield reference outside the sterile field in a closely correlated position and lighting conditions. Due to the significant curvature of the brain, this may not accurately account for the full vignetting profile so is not used for flatfield correction to produce quantitative data, and instead is used to generate spectral senstivities for relative data before demosaicing, cross-talk correction, and removal of spectra containing values above 99.5\% or below 0.5\%. These spectral sensitivities are found to be similar to those pre-calculated from the light source spectrum. This dataset is manually annotated using ImFusion Labels software to identify over 30 classes.

When fitting inverse models to these datasets we use a wavelength weighting similar to that seen in Chapter~\ref{chap:2layer} and reproduced in Figure~\ref{fig:weightings}, however analysis with uniform wavelength weighting can be seen in Appendix~\ref{ap:Chapter5uniform}. We also investigate two different fitting regimes where either the literature extinction coefficients for oxygenated and deoxygenated haemoglobin are used, or these extinction coefficients are shifted by 3nm before use. This 3nm shift was determined empirically as that which aligns with the peak positions seen in the HELICoiD mean annotated spectra. This difference in peak positions is likely driven by imperfect wavelength calibration in the HELICoiD system, however it could also arise from the difference in chemical environment of the haemoglobin in biological tissue compared to the acqueous configuration the literature values are measured in as mentioned for the dyes used in Chapter~\ref{chap:1layer}. The literature and shifted extinction coefficients can be seen in Figure~\ref{fig:extinctioncoeffs}.
% \begin{enumerate}
%     \item Wavelengths (450 - 600nm) are considered equally in the fitting of the residual. Literature extinction coefficients are used for oxyenated and deoxyenated haemoglobin. 
%     \item All wavelengths (450 - 600nm) are considered in the fitting of the residual, however those in the region of the haemoglobin peaks (525 - 585nm) are considered more heavily. Literature extinction coefficients are used for oxyenated and deoxyenated haemoglobin. 
%     \item Wavelengths (450 - 600nm) are considered equally in the fitting of the residual. Literature extinction coefficients for oxyenated and deoxyenated haemoglobin are shifted by 3nm before use. 
%     \item All wavelengths (450 - 600nm) are considered in the fitting of the residual, however those in the region of the haemoglobin peaks (525 - 585nm) are considered more heavily. Literature extinction coefficients for oxyenated and deoxyenated haemoglobin are shifted before use.
%     \label{ls:fittingmethods}
% \end{enumerate}
% The wavelength weightings and extinction coefficients are shown in Figure~\ref{fig:fittingoptions}. 
\begin{figure}[h!]
    \centering
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{fitting_region.pdf}
        \caption{}
        \label{fig:weightings}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{extinction_coefficients.pdf}
        \caption{}
        \label{fig:extinctioncoeffs}
    \end{subfigure}
    \caption{Wavelength weighting (left) and extinction coefficient (right) options used to fit models to neurosurgical HSI data.}
    \label{fig:fittingoptions}
\end{figure}

\subsection{Evaluation of analytical models}
The Yudovsky 2009 single and double layer models and Jacques 1999 models are evaluated as these are found to be the most effective in Chapters~\ref{chap:1layer} and \ref{chap:2layer}. These models are evaluated against datasets generated by simulating camera responses from the dermal and skin Monte Carlo datasets used in Chapters~\ref{chap:1layer} and \ref{chap:2layer} and so the diffuse reflectance prediction of these models are unchanged in this Chapter. When using these analytical models with the snapshot HSI datasets outlined in Sections~\ref{sec:MCcameras} to \ref{sec:NeuroHSIdata}, the predicted $R$ (detailed in Sections~\ref{sec:Jacques}, \ref{sec:Yudovskysingle}, and \ref{sec:Yudovsky2009}) is used to simulate noiseless camera responses as described by Equation~\eqref{eq:simulatingcam}. This simulation is not performed for the HELICoiD dataset as the camera band responses are not available in the dataset, however the authors report a very low dispersion of 0.74nm in the spectral range considered in this work so it is likely that the camera simulation would not have as significant of an impact with this system~\citep{Fabelo2018}. This allows the forwards and fitted models to be evaluated and quantified as described in Sections~\ref{sec:methodevaluate} and \ref{sec:methodevaluate2} against the Monte Carlo datasets, and similarly for the single-layer models for the gelatin-based phantom datsets.

For the gelatin-based phantom datasets, either the mean spectrum of the annotated region of a single frame is used, or each pixel in an annotated region from a single frame or from the temporal averaging of 5 frames is used. Both spatial and temporal averaging reduces the impact of noise on the resulting data. The mean annotated spectra are used for fitting the models (using the wavelength range 450-575nm) or to evaluate the quality of the forwards models with camera simulation. The quality of camera simulation can also be evaluated using $NRMSE$ and Pearson correlation coefficients by comparing simulated camera spectra from the spectrophotometer measurements of the gelatin-based phantoms in Section~\ref{sec:methodsphantoms}, to the measured HSI data in Section~\ref{sec:imagingphantoms}. The pixel-by-pixel analysis of the annotated region of either a single frame or the temporal average of 5 frames is used for fitting models to allow for spatially resolved results.

Finally, for each neurosurgical dataset image, the mean spectrum of each class annotation is calculated and the inverse analytical models are fitted to this (using the wavelength range 450-600cm). This provides a set of physiological parameters for each class of each image which are then collated to provide a range of values which are considered for plausibility. % Need to find better academic phrasing here

@@

\subsection{Monte Carlo}\label{sec:MCHSI}
The forwards analytical models are assessed using simulated camera responses generated from Monte Carlo data. 
%The Yudovsky 2009 and Jacques 1999 single-layer models are evaluated for three refractive indices, whereas the Yudovsky 2009 double-layer model is evaluated for $n=1.44$ only as in Chapters~\ref{chap:1layer} and~\ref{chap:2layer}. 
These are evaluated in Table~\ref{tb:forwardsHSIMC} for $n=1.44$ (further refractive indices can be found in Appendices~\ref{ap:forwardsHSIMCq} and \ref{ap:forwardsHSIMCr}). Examples of the quality of fitting for the single and double layer configurations with noiseless camera simulations are shown in Figure~\ref{fig:forwardsHSIMC}. Visual examples of the impact of noise on simulated MC spectra are seen in Figure~\ref{fig:forwardsMCnoise}.
\begin{table}[h!] %could remove noise levels here
    \centering
    \caption{Mean (standard deviation) $NRMSE$ (3.d.p.) between the simulated camera responses of each forwards spectrum from each model and each of 100 Monte Carlo simulated spectra using the same ground truth variable parameters for quantitative and relative data at a variety of noise levels (0, 1pp, 3pp). This is presented with the Pearson $r$ (bold if Pearson $p < 0.05$) for the linear regression between all forwards spectra against Monte Carlo simulated spectra for each refractive index dataset and each analytical model. All metrics are evaluated for the wavelength region of 450-600nm.}
    \begin{tabular}{|cc|c|ccc|ccc|}
        \hline
        Model & Layers & Quantitative & \multicolumn{3}{|c}{$NRMSE$} & \multicolumn{3}{|c|}{$r$} \\
         & & (Q) or & 0pp & 1pp & 3pp & 0pp & 1pp & 3pp \\
         & & Relative (R) &  &  &  &  & &  \\
        \hline
        \multirow{2}{*}{\shortstack{Yudovsky\\ 2009}} & \multirow{2}{*}{1} & Q & \shortstack{0.009 \\(0.004)} & \shortstack{0.082 \\(0.057)} & \shortstack{0.224 \\(0.125)} & \textbf{1.000} & \textbf{0.993} & \textbf{0.942} \\
        & & R & \shortstack{0.003 \\(0.002)} & \shortstack{0.075 \\(0.044)} & \shortstack{0.212 \\(0.118)} & \textbf{1.000} & \textbf{0.916} & \textbf{0.602} \\
        \hline
        \multirow{3}{*}{\shortstack{Jacques\\ 1999}} & \multirow{2}{*}{1} & Q & \shortstack{0.026 \\(0.046)} & \shortstack{0.089 \\(0.077)} & \shortstack{0.227 \\(0.130)} & \textbf{1.000} & \textbf{0.993} & \textbf{0.942} \\
        & & R & \shortstack{0.014 \\(0.021)} & \shortstack{0.076 \\(0.045)} & \shortstack{0.212 \\(0.117)} & \textbf{0.995} & \textbf{0.915} & \textbf{0.605} \\
        \hline
        \multirow{2}{*}{\shortstack{Yudovsky\\ 2009}} & \multirow{2}{*}{2} & Q & \shortstack{0.166 \\(0.133)} & \shortstack{0.482 \\(0.289)} & \shortstack{0.738 \\(0.294)} & \textbf{0.992} & \textbf{0.971} & \textbf{0.829} \\
         &  & R & \shortstack{0.033 \\(0.026)} & \shortstack{0.428 \\(0.286)} & \shortstack{0.684 \\(0.286)} & \textbf{0.991} & \textbf{0.270} & -0.007 \\
        \hline
    \end{tabular}
    \label{tb:forwardsHSIMC}
\end{table}

>>>>

\chapter[Application of models to HSI data]{Application of analytical tissue models to hyperspectral imaging data}\label{chap:HSImodel}
\begin{center}
\begin{minipage}[b]{0.9\linewidth}
\small
\textbf{Foreword\,}
This chapter is as yet unpublished. 
\newline
AB has conducted all of the research presented in this chapter under the supervision of JS, MB, and TV. The gelatin-based phantom images were annotated by OM. \footnote{\annref{18}\textcolor{red}{AB:~Anisha Bahl, JS:~Jonathan Shapey, MB:~Mads Bergholt, TV:~Tom Vercauteren, OM:~Oscar MacCormac}}
\end{minipage}
\end{center}

@@

\section{Introduction}
As described in Chapter~\ref{chap:intro}, hyperspectral imaging (HSI) is able to capture spatially resolved, diffuse reflectance spectra via non-contact imaging~\citep{Lu2014,Giannoni2018,Calin2014,Shapey2019}, which could enable quantitative, spatially-resolved, tissue oxygenation ($StO_2$) extraction. Many methods have been considered for this ranging from two or three wavelength approaches~\citep{MacKenzie2018} to wavelength-dependent analytical models including those investigated in Chapters~\ref{chap:1layer} and \ref{chap:2layer} as these have been some of the prominent methods of doing so~\citep{Yudovsky2009, Jacques1999, Clancy2015, Clancy2020}. Other methods include deep-learning based approaches, however these rely on models or simulation to generate their training data~\citep{Wirkert2016}, or models based on the derivatives of the spectra, however these rely on extensive calibration and assume that the scattering contribution is wavelength-independent which may not be true for all tissues~\citep{Holmer2018, Jacques2013}.

@@

In this chapter, we quantify the impact of reducing the spectral resolution and introduction of noise on the performance of the analytical models investigated in Chapters~\ref{chap:1layer} and \ref{chap:2layer} by simulating a snapshot camera response from the Monte Carlo datasets used in these previous chapters. These impacts are then evaluated with experimental HSI data of the phantoms detailed in Chapter~\ref{chap:1layer}. Finally, the models are applied to neurosurgical HSI data obtained intra-operatively as part of the HELICoiD dataset~\citep{Fabelo2019}. % and NHSI datasets.

\section{Methods}
\subsection{Camera response simulation}\label{sec:MCcameras}
To obtain datasets of simulated data with known ground truth parameters, the datasets described in Sections~\ref{sec:methodsMC} and \ref{sec:methodreference2} of internal and skin tissues are used. For each spectrum in the dataset, the camera response is simulated by convolving the band responses of the snapshot HSI camera (depicted in Figures~\ref{fig:rulerspectrum} and \ref{fig:simulatingcam}) and the integral of this is divided by the band response integral to account for differing band quantum efficiencies as defined in Equation~\eqref{eq:simulatingcam}. \annref{55}\textcolor{red}{By doing this, some key system parameters are introduced to adapt the datasets for snapshot HSI simulation.}
\begin{equation}
	c_{n} = \frac{\int r(\lambda)b_n(\lambda) d\lambda}{\int b_n(\lambda) d\lambda}
\label{eq:simulatingcam}
\end{equation}
Here $c_n$ is the measured value from band $n$ by the camera when the true spectrum is $r(\lambda)$ and the band response for this band is $b_n(\lambda)$. Each band response has a major peak around a central wavelength which is the wavelength associated with that band in the simulated spectrum. This is similar to the approach taken in Chapter~\ref{chap:SWB} for the ruler reflectivity correction as it takes into account both the true diffuse reflectance spectrum of the sample and the band responses of the camera used to capture it which do not perfectly sample single wavelengths. Random Gaussian noise can also be added at this stage where the mean is taken to be 0 and varying standard deviations ($\sigma$) are added. $c_n$ is then corrected using a correction matrix to account for the parasitic cross-talk effects as in Section~\ref{methodnecessity}~\citep{Pichette2017}. Whilst the illuminant spectrum will impact the measured spectrum, we employ appropriate white balancing as described in Chapter~\ref{chap:SWB} which should limit the impact of this as well as compensate for the reflectivity of the white reference so the illuminant spectrum is not included in this simulation. \selfref{11}\textcolor{red}{Additionally, the optical transmission of other optical components, such as the coupler or exoscope, in the system should be accounted for which are assumed to be perfectly transparent to all wavelengths in this simulation.} An example of the stages of this process can be seen in Figure~\ref{fig:simulatingcam}. %This allows a variety of datasets to be simulated with the same set of ground truth parameters for both quantitative and mean normalised relative data. 
Two noise levels are investigated alongside noiseless simulations. The maximum noise level chosen ($\sigma = 3pp$ before mean normalisation) is similar to the standard deviation size seen in the highly controlled dataset investigated in Chapter~\ref{chap:SWB} in units of percentage point (pp).  
\begin{figure}[h!]
    \centering 
    \begin{subfigure}{0.7\textwidth}
    \centering
        \includegraphics[width=\textwidth]{simulatingcam.pdf}
        \caption{}
        \label{fig:simulatingcamsubfigure}
    \end{subfigure}
    \begin{subfigure}{0.62\textwidth}
        \centering
        \includegraphics[width=\textwidth]{band_responses.pdf}
        \caption{}
        \label{fig:bandresponses}
    \end{subfigure}
    % \includegraphics[width=0.7\textwidth]{simulatingcam.pdf}
    \caption{Top: An example of the stages of converting a Monte Carlo diffuse reflectance spectrum (\textcolor{blue}{blue solid}) to a simulated camera response (\textcolor{red}{red $*$}) via the band response convolution (\textcolor{purple}{purple $+$}) and cross-talk correction. Bottom: A depiction of the band responses used in this camera simulation.}
    \label{fig:simulatingcam}
\end{figure}

@@

    \caption{Table displaying the percentages of dyes [acid red 1 (AR1), acid red 14 (AR14), crystal violet (CV)] used for each dye configuration alongside the total dye scaled concentration in arbitrary units. The intralipid (v/v \%) imaged for each configuration are also indicated.}

@@

A white reference of a Spectralon tile with 95\% reflectivity is taken under the same lighting conditions and imaging plane as the phantoms. This white reference is either used directly for white balancing a frame of the hyperspectral video to produce quantitative data or to generate spectral sensitivities used to construct mean normalised relative data as in Chapter~\ref{chap:SWB}. This demonstrates both the highest fidelity and more realistic data processing pathways. Following this, bilinear demosaicing is used and cross-talk correction is applied per pixel of the frame also detailed in Chapter~\ref{chap:SWB}. At this stage, it is possible to temporally average frames to improve the signal to noise ratio of the resulting hypercube. Finally, any pixel spectra containing values above 99.5\% or below 0.5\% are discarded to reduce the impact of specular reflections and to ensure the data is within the dynamic range of the sensor. 

@@

\subsection{Neurosurgical HSI dataset}\label{sec:NeuroHSIdata}
The HELICoiD dataset is leveraged in this work which captures data in the wavelength range 400-1000nm using a pushbroom spatial scanning HSI camera~\citep{Fabelo2019}. This data is obtained as part of a collaboration between four universities, three industrial partners, and two hospitals in the European Union~\citep{Fabelo2019}. This provides 36 high spectral resolution images of 22 patients which are annotated with use of spectral angle mapper to identify four key classes: normal tissue, tumour tissue, hypervascularised tissue, and background elements. This data is white balanced using associated flatfield white references included in the dataset. The resulting spectra are then normalised to reduce the impact of any differences in relative position of the camera to the data and white reference as discussed in Chapter~\ref{chap:SWB}. For this work we consider only the spectral range of 450-600nm as this is the range of use for the analytical models considered. 
%Secondly, a NeuroHSI dataset is considered which is obtained using the same snapshot mosaic camera (Ximea utilising the IMEC CMV2K-SSM4X4-VIS sensor, Germany) as in Section~\ref{sec:imagingphantoms} whose bands are within the 450-600nm wavelength range. In contrast to Section~\ref{sec:imagingphantoms}, this is mounted to a neurosurgical microscope (Zeiss Kinevo 900, Germany) whose integrated light source and lenses are used for simpler integration into the neurosurgical workflow, allowing for use in a wide variety of neurosurgical cases. 
%%This light source and optical system was not fully characterised and therefore the vignetting profile cannot be assumed to follow the trends seen in Chapter~\ref{chap:SWB} and so the full white balancing algorithm using the sterile ruler cannot be implemented for quantitative data. In it's place, 
%In each imaging timepoint of this study, the position of the neurosurgical microscope is recorded and used to obtain a flatfield reference outside the sterile field in a closely correlated position and lighting conditions. Due to the significant curvature of the brain, this may not accurately account for the full vignetting profile so is not used for flatfield correction to produce quantitative data, and instead is used to generate spectral senstivities for relative data before demosaicing, cross-talk correction, and removal of spectra containing values above 99.5\% or below 0.5\%. These spectral sensitivities are found to be similar to those pre-calculated from the light source spectrum. This dataset is manually annotated using ImFusion Labels software to identify over 30 classes.

When fitting inverse models to these datasets we use a wavelength weighting similar to that seen in Chapter~\ref{chap:2layer} and reproduced in Figure~\ref{fig:weightings}, however analysis with uniform wavelength weighting can be seen in Appendix~\ref{ap:Chapter5uniform}. We also investigate two different fitting regimes where either the literature extinction coefficients for oxygenated and deoxygenated haemoglobin are used, or these extinction coefficients are shifted by 3nm before use. This \selfref{12}\textcolor{red}{simplistic} 3nm shift was determined empirically as that which aligns with the peak positions seen in the HELICoiD mean annotated spectra. This difference in peak positions is likely driven by imperfect wavelength calibration in the HELICoiD system, however it could also arise from the difference in chemical environment of the haemoglobin in biological tissue compared to the aqueous configuration in which the literature values are measured. The latter effect could be similar to that seen in the dyes used in Chapter~\ref{chap:1layer}\selfref{12}\textcolor{red}{, meaning that other transformations could be necessary such as peak broadening}. The literature and shifted extinction coefficients can be seen in Figure~\ref{fig:extinctioncoeffs}.
% \begin{enumerate}
%     \item Wavelengths (450 - 600nm) are considered equally in the fitting of the residual. Literature extinction coefficients are used for oxyenated and deoxyenated haemoglobin. 
%     \item All wavelengths (450 - 600nm) are considered in the fitting of the residual, however those in the region of the haemoglobin peaks (525 - 585nm) are considered more heavily. Literature extinction coefficients are used for oxyenated and deoxyenated haemoglobin. 
%     \item Wavelengths (450 - 600nm) are considered equally in the fitting of the residual. Literature extinction coefficients for oxyenated and deoxyenated haemoglobin are shifted by 3nm before use. 
%     \item All wavelengths (450 - 600nm) are considered in the fitting of the residual, however those in the region of the haemoglobin peaks (525 - 585nm) are considered more heavily. Literature extinction coefficients for oxyenated and deoxyenated haemoglobin are shifted before use.
%     \label{ls:fittingmethods}
% \end{enumerate}
% The wavelength weightings and extinction coefficients are shown in Figure~\ref{fig:fittingoptions}. 
\begin{figure}[h!]
    \centering
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{fitting_region.pdf}
        \caption{}
        \label{fig:weightings}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{extinction_coefficients.pdf}
        \caption{}
        \label{fig:extinctioncoeffs}
    \end{subfigure}
    \caption{The wavelength weighting (left) and extinction coefficient options (right) used to fit models to neurosurgical HSI data.}
    \label{fig:fittingoptions}
\end{figure}

\subsection{Evaluation of analytical models}
The Yudovsky 2009 single and double-layer models and Jacques 1999 models are evaluated as these are found to be the most effective in Chapters~\ref{chap:1layer} and \ref{chap:2layer}. These models are evaluated against datasets generated by simulating camera responses from the dermal and skin Monte Carlo datasets used in Chapters~\ref{chap:1layer} and \ref{chap:2layer} and so the diffuse reflectance prediction of these models are unchanged in this Chapter. When using these analytical models with the snapshot HSI datasets outlined in Sections~\ref{sec:MCcameras} to \ref{sec:NeuroHSIdata}, the predicted $R$ (detailed in Sections~\ref{sec:Jacques}, \ref{sec:Yudovskysingle}, and \ref{sec:Yudovsky2009}) is used to simulate noiseless camera responses as described by Equation~\eqref{eq:simulatingcam}. This allows the forwards and fitted models to be evaluated and quantified as described in Sections~\ref{sec:methodevaluate} and \ref{sec:methodevaluate2} against the Monte Carlo datasets, and similarly the single-layer models can be evaluated using the gelatin-based phantom datsets.

For the gelatin-based phantom datasets, either the mean spectrum of the annotated region of a single frame is used, or each pixel in an annotated region from a single frame or from the temporal averaging of 5 frames is used. Both spatial and temporal averaging reduces the impact of noise on the resulting data. The mean annotated spectra are used for fitting the models (using the wavelength range 450-575nm) or to evaluate the quality of the forwards models with camera simulation. The quality of camera simulation can also be evaluated %using $NRMSE$ and Pearson correlation coefficients 
by comparing simulated camera spectra from the spectrophotometer measurements of the gelatin-based phantoms in Section~\ref{sec:methodsphantoms}, to the measured HSI data in Section~\ref{sec:imagingphantoms}. The pixel-by-pixel analysis of the annotated region of either a single frame or the temporal average of 5 frames is used for fitting models to allow for spatially resolved results. All evaluation parameters used for this replicate those detailed in Sections~\ref{sec:methodevaluate} and \ref{sec:methodevaluate2}.

Finally, for each neurosurgical dataset image, the mean spectrum of each class annotation is calculated and the inverse analytical models are fitted to this (using the wavelength range 450-600nm). This provides a set of physiological parameters for each class of each image which are then collated to provide a range of $StO_2$ values which are compared to typical jugular venous oxygen saturation (JVOS) and cerebral oximetry values. The camera simulation is not performed when fitting models to the HELICoiD dataset as the camera band responses are not available in the dataset, however the authors report a very low dispersion of 0.74nm in the spectral range considered in this work so it is likely that the camera simulation would not have as significant of an impact with this system~\citep{Fabelo2018}. % Need to find better academic phrasing here

@@

\subsection{Monte Carlo}\label{sec:MCHSI}
The forwards analytical models are assessed using simulated camera responses generated from Monte Carlo data. 
%The Yudovsky 2009 and Jacques 1999 single-layer models are evaluated for three refractive indices, whereas the Yudovsky 2009 double-layer model is evaluated for $n=1.44$ only as in Chapters~\ref{chap:1layer} and~\ref{chap:2layer}. 
These are evaluated in Table~\ref{tb:forwardsHSIMC} for $n=1.44$ (further refractive indices can be found in Appendices~\ref{ap:forwardsHSIMCq} and \ref{ap:forwardsHSIMCr}). Examples of the quality of fitting for the single and double-layer configurations with noiseless camera simulations are shown in Figure~\ref{fig:forwardsHSIMC}. Visual examples of the impact of noise on simulated MC spectra are seen in Figure~\ref{fig:forwardsMCnoise}.
\begin{table}[h!] %could remove noise levels here
    \centering
    \caption{Mean (standard deviation) $NRMSE$ (3.d.p.) between the simulated camera responses of each forwards spectrum from each model and each of 100 Monte Carlo simulated spectra using the same ground truth variable parameters for quantitative (Q) and relative (R) data at a variety of noise levels (0pp, 1pp, 3pp). This is presented with the Pearson $r$ (bold if $p < 0.05$) for the linear regression between all forwards spectra against Monte Carlo simulated spectra for each refractive index dataset and each analytical model. All metrics are evaluated for the wavelength region of 450-600nm.}
    \begin{tabular}{|cc|c|ccc|ccc|}
        \hline
        Model & Layers & Quantitative & \multicolumn{3}{|c}{$NRMSE$} & \multicolumn{3}{|c|}{$r$} \\
         & & (Q) or & 0pp & 1pp & 3pp & 0pp & 1pp & 3pp \\
         & & Relative (R) &  &  &  &  & &  \\
        \hline
        \multirow{2}{*}{\shortstack{Yudovsky\\ 2009}} & \multirow{2}{*}{1} & Q & \shortstack{0.009 \\(0.004)} & \shortstack{0.082 \\(0.057)} & \shortstack{0.224 \\(0.125)} & \textbf{1.000} & \textbf{0.993} & \textbf{0.942} \\
        & & R & \shortstack{0.003 \\(0.002)} & \shortstack{0.075 \\(0.044)} & \shortstack{0.212 \\(0.118)} & \textbf{1.000} & \textbf{0.916} & \textbf{0.602} \\
        \hline
        \multirow{3}{*}{\shortstack{Jacques\\ 1999}} & \multirow{2}{*}{1} & Q & \shortstack{0.026 \\(0.046)} & \shortstack{0.089 \\(0.077)} & \shortstack{0.227 \\(0.130)} & \textbf{1.000} & \textbf{0.993} & \textbf{0.942} \\
        & & R & \shortstack{0.014 \\(0.021)} & \shortstack{0.076 \\(0.045)} & \shortstack{0.212 \\(0.117)} & \textbf{0.995} & \textbf{0.915} & \textbf{0.605} \\
        \hline
        \multirow{2}{*}{\shortstack{Yudovsky\\ 2009}} & \multirow{2}{*}{2} & Q & \shortstack{0.166 \\(0.133)} & \shortstack{0.482 \\(0.289)} & \shortstack{0.738 \\(0.294)} & \textbf{0.992} & \textbf{0.971} & \textbf{0.829} \\
         &  & R & \shortstack{0.033 \\(0.026)} & \shortstack{0.428 \\(0.286)} & \shortstack{0.684 \\(0.286)} & \textbf{0.991} & \textbf{0.270} & -0.007 \\
        \hline
    \end{tabular}
    \label{tb:forwardsHSIMC}
\end{table}
